---
title: "Parallel computation and reproducibility"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parallel computation and reproducibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(cocorresp)
```

`crossval()` and `vegan::permutest()` run serially by default. Leave-one-out folds
are independent; permutation replicates within an axis are independent, but
successive axes depend on residualisation and must run sequentially.

```{r data}
data(beetles)
data(plants)
y <- log1p(beetles)
x <- plants
fit <- coca(y, x, n.axes = 2, quiet = TRUE)
set.seed(2026)
vegan::permutest(fit, permutations = 9, verbose = FALSE)
```

## Worker ownership

Pass `parallel = 2L` to create two PSOCK workers, automatically stopped on
success or error. Pass an existing cluster to reuse workers; cocorresp never
stops a caller-owned cluster. Avoid nested parallelism and limit BLAS threads
when each worker performs matrix operations. Small datasets often run faster
serially because worker startup and serialisation have a cost.

```{r clusters, eval=FALSE}
crossval(y, x, n.axes = 2, parallel = 2L)
local({
  cl <- parallel::makePSOCKcluster(2)
  on.exit(parallel::stopCluster(cl))
  vegan::permutest(fit, parallel = cl)
})
local({
  cl <- mirai::make_cluster(2)
  on.exit(mirai::stop_cluster(cl))
  crossval(y, x, n.axes = 2, parallel = cl)
})
```

## Future and futurize adapters

A mapping function must accept `X`, `FUN`, and `...`, and return a list in
input order. The package checks this contract. Backend packages remain
optional. Install the backend of your choice separately.

```{r futures, eval=FALSE}
future_map <- function(X, FUN, ...) {
  future.apply::future_lapply(X, FUN, ..., future.seed = FALSE)
}
futurize_map <- function(X, FUN, ...) {
  futurize::futurize(lapply(X, FUN, ...), seed = FALSE)
}
local({
  old <- future::plan(future::multisession, workers = 2)
  on.exit(future::plan(old))
  crossval(y, x, n.axes = 2, parallel = future_map)
  vegan::permutest(fit, parallel = futurize_map)
})
# Alternatively select future.mirai::mirai_multisession in future::plan().
```

Wrapping `crossval(...)` directly in `futurize()` is not a supported substitute
for the adapter. cocorresp never selects a future plan or launches mirai daemons
unless the user supplies an appropriate backend.

## Random numbers and restricted permutations

All permutations are generated in the calling process. Workers perform no
random draws. For a fixed seed, RNG configuration, and input, changing backend
or worker count preserves permutation statistics and the resulting RNG state.
Floating-point results can vary slightly across BLAS implementations.

Integer counts preserve the historical algorithm: a fresh set of permutations
is sampled for every axis, with residuals assigned into sampled rows. Explicit
matrices instead select residual rows directly, as in `residuals[p, ]`.
A control object generates one matrix reused across axes. Consequently, a
control and a count need not produce identical results from the same seed.

```{r designs}
control <- permute::how(nperm = 9, blocks = rep(1:3, each = 10))
set.seed(2026)
pm <- permute::shuffleSet(nrow(y), control = control)
vegan::permutest(fit, permutations = pm, verbose = FALSE)
```

Rows of a permutation matrix must contain each observation index exactly once.
The number of matrix rows determines the number of replicates. The observed
statistic is included once in the p-value calculation, `(1 + exceedances)/(1 + B)`.
Small or restrictive designs can yield fewer permutations than requested;
the actual count is used. Users are responsible for exchangeability assumptions.

For `vegan::envfit(..., na.rm = TRUE)`, free blocked designs and strata are reduced
alongside observations. Explicit matrices must preserve the retained subset.
For series, grid, or plot designs, construct complete-case data and a matching
control explicitly; the package errors rather than silently changing the design.
